Improving the portuguese hyphenation rules

### 2023-07-18 article/article.tex
fix #27 - table still needs update and would fit better as an attachment
text update

In some situations, hyphenation is also a matter of style. Some partitioning
choices sound better than others. These conflicting alternatives typically
arises when a words has many possible hyphenation points. Consider
\emph{ar-chae-ol-o-gist} (\emph{or ar-che-ol-o-gist}), which is preferably partitioned as
\emph{archae-ologist} (or \emph{arche-ologist}) in opposition to \emph{archaeol-ogist} (or
\emph{archeol-ogist}) or \emph{archaeolo-gist} (or \emph{archeolo-gist}). It is preferable to keep
whole morphemes together. In the previous example: \emph{archae} (or \emph{arche}, meaning
``ancient'', ``primitive'')
%\footnote{archae- comes from the Ancient Greek ἀρχαῖος (arkhaîos, `ancient',
%`primitive'), from ἀρχή (arkhḗ, `beginning').}
%https://en.wiktionary.org/wiki/archae-
and \emph{-ologist} (``one who studies the topic''). In Portuguese, it is also more
elegant to avoid splits between double consonants or vowels, even if an
hyphenation point do exists between those letters. For example,
\emph{pressu-rizar} is preferable over \emph{pres-surizar} and
\emph{empreen-dedor} is preferable over \emph{empre-endedor}. Even so,
exceptions exists, it is preferable to partition \emph{micro-organismo}
rather than \emph{microor-ganismo} (keeping morphemes together is favored
over splitting a double vowel).

The general rule for Portuguese hyphenation is to split a word into its
syllables. A syllable is made of a mandatory nucleus, filled by a vowel, and
optional peripheral consonants (before or after the nucleus). In some
situations, the syllabic division does not respect the ethnologic constituents.
The usage of the prefixes \emph{bis-} and \emph{in-} are examples of this
circumstance. The correct syllabifications are \emph{bi-sa-vô},
\emph{i-na-ti-vo}, and \emph{i-nobs-tan-te}\footnote{The rule of syllable
division could lead to two possible partitions: \emph{i-nobs-tan-te} and
\emph{in-obs-tan-te}, but the first is preferable.}, where the prefixes are
split into two syllables. But, as pointed previously, it is preferable to keep
morphemes together rather than splitting them apart, therefore the we should
favor the hyphenation \emph{bisa-vô} over \emph{bi-savô} and \emph{ina-tivo}
over \emph{i-nativo}. This last example could also lead to a misunderstanding
since the word \emph{nativo} (\emph{native} in English) emerges from this word
break.

@@

Machine learning approach was used by to hyphenate Norwegian text
\cite{kristensen2001}. It was found that the \TeX{} approach, in general,
performed better then a neural network. Both approaches had similar
performance on finding the correct hyphenation points and avoiding wrong
hyphenation points on a small word list (the neural network performed slightly
better on finding the correct hyphenation points). Nonetheless, \TeX{} approach overcame
the neural network on a large word list, regarding the avoidance of wrong
hyphenations.

The original \TeX{} hyphenation algorithm was introduced by Knuth 1977
\cite{knuth1977}. It was focused on the English language and used three main
rules:(1) suffix removal; (2) prefix removal; and (3)
vowel-consonant-consonant-vowel (VCCV) breaking. Tests showed it could find
40\% of the allowed hyphen locations \cite{liang1983}. The hyphenation
algorithm proposed by Frank M. Liang and adopted in \TeX{} uses the notion of
competing patterns. A database of hyphenated words is swept, looking for
hyphenating and inhibiting patterns. The algorithm introduced in \TeX{}82 uses
five alternating levels of hyphenating and inhibiting patterns. The program
for patten generation, based on a corpus, PATGEN was created Liang
\cite{liangbreitenlohner1999} and was used to create hyphenating patterns for
many languages
\cite{sojka1995,sojka1995a,sojka2005thesis,sojka2003,scannell2003}.

The effective hyphenation of words by \TeX{} will actually depends on the following factors:
\begin{enumerate*}[label=\arabic*)]
    \item document language, which will determine which set of patterns to apply;
    \item characters used, since some might block hyphenation at their edges;
    \item the value of the internal variables \verb|lefthyphenmin| and \verb|righthyphenmin|\footnote{
	The variables \verb|lefthyphenmin| and \verb|righthyphenmin| are language dependant and
	are defined in \emph{tlpobj} files (\verb|/usr/local/texlive/20XX/tlpkg/tlpobj/hyphen-xxxxxx.tlpobj|). 
	Default values varies in the range from 1 to 3. 
    	English and Portuguese, for example, use \verb|lefthyphenmin=2| and \verb|righthyphenmin=3|.},
	% ls /usr/local/texlive/2023/tlpkg/tlpobj/*.tlpobj | while read -r filename; do grep -Po "lefthyphenmin=[0-9]\s+righthyphenmin=[0-9]" $filename; done | tr '\t' ' ' | tr -s ' ' | sort | uniq -c
	%      17 lefthyphenmin=1 righthyphenmin=1
        %       6 lefthyphenmin=1 righthyphenmin=2
        %       1 lefthyphenmin=1 righthyphenmin=3
        %      49 lefthyphenmin=2 righthyphenmin=2
        %       9 lefthyphenmin=2 righthyphenmin=3
        which defines the minimum sequence length of characters at the left and right borders
        before any hyphenation is allowed.
\end{enumerate*}

Despite the fact that \TeX{} hyphenation algorithm and rules are old, they are,
to these days, the most frequently used even outside the \TeX{}'s world. The
grounds for this is Hunspell, a spell checker and morphological analyzer that
is adopted in may software (e.g. LibreOffice, OpenOffice.org, Mozilla Firefox,
Mozilla Thunderbird, Google Chrome, macOS, InDesign, memoQ, Opera, Affinity
Publisher, among others \cite{hunspell}). Hunspell uses \TeX{} hyphenation
rules \cite{hunspellhyphen,levien1998}, making \TeX{} hyphenation widespread in
the computer world. That is a result of \TeX{} approach simplicity and
versatility.  The algorithm works well, it currently has rules for 66 languages
\cite{texhyphenrules} and it is possible to design new rules for those still
missing.

Unfortunately, some specific hyphenation rules are not possible to be
implemented using the \TeX{} hyphenation algorithm. In German, for example,
when words are hyphenated, some letters might change or a letter might be
inserted.  Compound words have no hyphens, leading to long sequences of letters
with no visible separation and even repetitions of a same letter, such as in
\emph{Wasserrinne} and \emph{Schifffahrt}. Furthermore, the German spelling
reform made some changes, making it necessary to create a different set of
rules for German hyphenation.  For example, the word \emph{Schiffahrt} should
be hyphenated as \emph{Schiff-fahrt}, preserving the \emph{f}s from each word
that makes this compound. The hyphenation should insert a \emph{f} that is not
part of the written form. That was not a problem for the old written form of
the word: \emph{Schifffahrt}. Also, the old hyphenating rules of German grammar
stated the hyphenation \emph{Bäk-ker} for the word \emph{Bäcker},
\emph{Zuk-ker} for the word \emph{Zucker} and \emph{pak-ken} for the word
\emph{packen}. Now those words are hyphenated as \emph{Bä-cker}, \emph{Zu-cker}
and \emph{pa-cken}, respectively. Some words have also changed their
hyphenation point after the spelling reform. For example, \emph{Fen-ster}
became \emph{Fens-ter} and \emph{mei-stens} became \emph{meis-tens}. Some
problems in compound word hyphenation in \TeX{} are discussed in
\cite{sojka1995a}.

@@

\section{PATGEN}\label{sec-patgen}
PATGEN uses a list of hyphenated words to define rules at different
levels and lengths, based on the patterns found in the data. It start with
short patterns and increases until a maximum pattern length, allowed by the
user, is reached. As it advances, creating longer patterns, it also establishes
exceptions. In some cases, the analysis of long patterns might be necessary,
since some hyphenation points might depend on characters far away from the
breaking point\footnote{Some examples of hyphenation dependency on characters
far from the break point: \emph{dem-o-crat} and \emph{de-moc-ra-cy};
\emph{as-pi-rin} and \emph{aspir-ing}; \emph{de-mon-stra-tive} and
\emph{dem-on-stra-tion}.}.

PATEGEN works on glyph indices rather than character codes. Each glyph is
represented by a single byte. That amount to 256 indices, where 13 of them are
reserved for the digits 0-9 and the characters `.', `-' and `*'. The remaining 243
are used to represent symbols of the given language. To run PATGEN, it is required a
translation file, which defines the values of some language specific parameters
(in the first line) and defines the many forms in which language symbols might
appear (all following lines). In the first line, positions 1 and 2 are used to
set the value of \verb|lefthyphenmin|, positions 3 and 4 are for
\verb|righthyphenmin|. These values defines the shortest length of a string
that might be generated by an hyphenation procedure. In order to set a single
digit value, leave the first position blank, that is, put a blank space in
position 1 and 3 for \verb|lefthyphenmin| and \verb|righthyphenmin|
respectively. Positions 5, 6 and 7 are used to define alternative values for
the special characters `.', `-' and `*'. The following lines uses a delimiter
to surround each `letter' of the desired language alphabet and also its
alternative representations. The first position of the line defines the
delimiter and each symbol of the language may use as many positions as
necessary, since not using the value reserved for the delimiter. Consider the
example for defining the letter `e' in Portuguese: 
\begin{verbatim}
XeXEXéXêXÉXÊX\'{e}X\^{e}X\'{E}X\^{E}XX 
\end{verbatim}
It represents the many ways in which it might be found: lowercase, uppercase,
with or without acute or circumflex accents. In this definition, we have
assumed that the many forms in which we may find the character \emph{e} and
\emph{e} will be equivalent for hyphenation (pattern matching) purpose. As
another example, see the line defining the character $\pi$ (taken from
\cite{haralambous2021}): \begin{verbatim} #p#P#\varpi ## \end{verbatim}

The translate file and the dictionary file should use the same encoding. Even
if it is a multi byte encoding, the translation file express how to deal with a
sequence of bytes representing a glyph and PATGEN will work fine as long as
there are at most 243 symbols in the given language.
The dictionary file is a list of already hyphenated words from which PATGEN
will look for patterns and create hyphenation rules.

Those rules are organized
in different levels, from 1 to 9, where odd numbers represent hyphenating
levels and even numbers represent inhibiting levels. Each level works as an
inhibiting level of it predecessor. 
%A pattern will consist of a string made of
%characters (from the language alphabet) possibly with a number in between,
%expressing the hyphenation/inhibition level and possibly word boundaries marker
%(the period) at the pattern edges. When there is no number between characters
%in a pattern, a zero is assumed, which means \emph{undefined} and no
%hyphenation point will be suggest at that location.

\section{\TeX{} hyphenation rules for portuguese}\label{sec-tex-hyphen-pt}

\cite{rezende1987} created the first patters for Portuguese hyphenation in
\TeX{}. It was lastly updated in 2015 with the contributions from José João
Dias Almeida \cite{hyphpt}. The set has \NumberOfDefaultRules{}
rules\footnote{The set of rules for Portuguese hyphenation is short when
compared to rules in other languages. For example, English currently has 4938
rules, Russian has 7023 rules and German has 34011 rules.} and are able to
correctly hyphenate the majority of Portuguese words. 
% - detalhar as regras de Rezende e Dias 
% - contabilizar a quantidade de palavras que não são contempladas pela proposta do autor
% - analisar a necessidade de excluir alguma regra (ao invés de apenas acrescentar novas)

In front of these cases, we propose here to analyse the performance of the default rules in order to add new rules to improve
hyphenation by correcting some issues and taking in account non-typical
patterns. We describe the methodology used in the section \ref{sec-methodology}.

\section{Methodology}\label{sec-methodology}

In order to apprise the performance of each set of rules, we need a list of
correctly hyphenated words that are used in Portuguese. For that, we use the
frequency of occurrence to avoid selecting really rare words. To build a large
list with such information, we relied on data collected from online
dictionaries and corpora of Portuguese. We also looked up Portuguese grammars
to check on the hyphenation (\emph{translineation}) rules and we made some manual
corrections on the data.

@@

Initially we selected \href{https://www.linguateca.pt/cetenfolha/index_info.html}{CETENFolha} 
as the source corpus. Using only this corpus led to two problems: many words
were still missing and it was built on text from 1994, prior to the
Orthographic Agreement. This problem is a result of the Orthographic Agreement
of the Portuguese Language, carried out in 1990.

The Agreement's adopting transition period started in 2009 and ended as it
turned mandatory in 2016.  Even after the agreement, some words still hold
different spellings in the participant countries. It is desirable to have
hyphenation rules that might work properly, despite those idiosyncrasies.  For
example, the word \emph{reception} might be written as \emph{receção}
(Portugal) or \emph{recepção} (Brazil); the word \emph{action} might be written
as \emph{acção} (Portugal) or \emph{ação} (Brazil); and the word \emph{project}
might be written as \emph{projecto} (Portugal) or \emph{projeto} (Brazil).

@@

O conjunto de \NumberOfNewRules{} regras é apresentado a seguir:
\begin{multicols}{4}
    \verbatiminput{patch.TeX.pt-br.patterns}
\end{multicols}

>>>>

In some situations, hyphenation is also a matter of style. Some partitioning
choices sound better than others. These conflicting alternatives typically
arises when a words has many possible hyphenation points. Consider
\emph{ar-chae-ol-o-gist} (\emph{or ar-che-ol-o-gist}), which is preferably partitioned as
\emph{archae-ologist} (or \emph{arche-ologist}) in opposition to \emph{archaeol-ogist} (or
\emph{archeol-ogist}) or \emph{archaeolo-gist} (or \emph{archeolo-gist}). It is preferable to keep
whole morphemes together. In the previous example: \emph{archae} (or \emph{arche}, meaning
``ancient'', ``primitive'')
%\footnote{archae- comes from the Ancient Greek ἀρχαῖος (arkhaîos, `ancient',
%`primitive'), from ἀρχή (arkhḗ, `beginning').}
%https://en.wiktionary.org/wiki/archae-
and \emph{-ologist} (``one who studies the topic''). In Portuguese, it is also more
elegant to avoid splits between double consonants or vowels, even if an
hyphenation point do exists between those letters. For example,
\emph{pressu-rizar} is preferable over \emph{pres-surizar} and
\emph{empreen-dedor} is preferable over \emph{empre-endedor}. Even so,
exceptions exists, it is preferable to partition \emph{micro-organismo}
rather than \emph{microor-ganismo} (keeping morphemes together is favored
over splitting a double vowel). Numerous factors come into play when choosing
the optimal hyphenation point for a word.

The general rule for Portuguese hyphenation is to split a word into its
syllables. A syllable is made of a mandatory nucleus, filled by a vowel, and
optional peripheral consonants (before or after the nucleus). In some
situations, the syllabic division does not respect the ethnologic constituents.
The usage of the prefixes \emph{bis-} and \emph{in-} are examples of this
circumstance. The correct syllabifications are \emph{bi-sa-vô},
\emph{i-na-ti-vo}, and \emph{i-nobs-tan-te}\footnote{The rule of syllable
division could lead to two possible partitions: \emph{i-nobs-tan-te} and
\emph{in-obs-tan-te}, but the first is preferable.}, where the prefixes are
split into two syllables. But, as pointed previously, it is preferable to keep
morphemes together rather than splitting them apart, therefore we should
favor the hyphenation \emph{bisa-vô} over \emph{bi-savô}, and \emph{ina-tivo}
over \emph{i-nativo}. This last example could also lead to a misunderstanding
since the word \emph{nativo} (\emph{native} in English) emerges from this word
break.

@@

Machine learning approach was used to hyphenate Norwegian text
\cite{kristensen2001}. It was found that the \TeX{} approach, in general,
performed better then a neural network. Both approaches had similar
performance on finding the correct hyphenation points and avoiding wrong
hyphenation points on a small word list (the neural network performed slightly
better on finding the correct hyphenation points). Nonetheless, \TeX{} approach overcame
the neural network on a large word list, regarding the avoidance of wrong
hyphenations.

The original \TeX{} hyphenation algorithm was introduced by Knuth 1977
\cite{knuth1977}. It was focused on the English language and used three main
rules:
\begin{enumerate*}[label=(\arabic*)]
 \item suffix removal; 
 \item prefix removal; and 
 \item vowel-consonant-consonant-vowel (VCCV) breaking. 
\end{enumerate*}
Tests showed it could find
40\% of the allowed hyphen locations \cite{liang1983}. 
The hyphenation algorithm proposed by Frank M. Liang and adopted in \TeX{}
utilizes competing patterns. The \TeX{}82 algorithm employs five alternating
levels of hyphenating and inhibiting patterns. The program for pattern
generation, known as PATGEN, was created by Liang \cite{liangbreitenlohner1999}
and has been utilized to generate hyphenating patterns for numerous languages
\cite{sojka1995,sojka1995a,sojka2005thesis,sojka2003,scannell2003}. It involves
sweeping a database of hyphenated words in a language to identify both
hyphenating and inhibiting patterns, ultimately creating a list of competing
patterns for that specific language.

The effective hyphenation of words by \TeX{} will actually depends on the following factors:
\begin{enumerate*}[label=(\arabic*)]
    \item document language, which will determine which set of patterns to apply;
    \item characters used, since some might block hyphenation at their edges;
    \item the value of the internal variables \verb|lefthyphenmin| and \verb|righthyphenmin|\footnote{
	The variables \verb|lefthyphenmin| and \verb|righthyphenmin| are language dependant and
	are defined in \emph{tlpobj} files (\verb|/usr/local/texlive/20XX/tlpkg/tlpobj/hyphen-xxxxxx.tlpobj|). 
	Default values varies in the range from 1 to 3. 
    	English and Portuguese, for example, use \verb|lefthyphenmin=2| and \verb|righthyphenmin=3|.},
	% ls /usr/local/texlive/2023/tlpkg/tlpobj/*.tlpobj | while read -r filename; do grep -Po "lefthyphenmin=[0-9]\s+righthyphenmin=[0-9]" $filename; done | tr '\t' ' ' | tr -s ' ' | sort | uniq -c
	%      17 lefthyphenmin=1 righthyphenmin=1
        %       6 lefthyphenmin=1 righthyphenmin=2
        %       1 lefthyphenmin=1 righthyphenmin=3
        %      49 lefthyphenmin=2 righthyphenmin=2
        %       9 lefthyphenmin=2 righthyphenmin=3
        which defines the minimum sequence length of characters at the left and right borders
        before any hyphenation is allowed.
\end{enumerate*}

Despite the fact that \TeX{} hyphenation algorithm and rules are old, they are,
to these days, the most frequently used approach, even outside the \TeX{}'s
world. The grounds for this is Hunspell, a spell checker and morphological
analyzer that is adopted in many softwares (e.g. LibreOffice, OpenOffice.org,
Mozilla Firefox, Mozilla Thunderbird, Google Chrome, macOS, InDesign, memoQ,
Opera, Affinity Publisher, among others \cite{hunspell}). Hunspell uses \TeX{}
hyphenation rules \cite{hunspellhyphen,levien1998}, making \TeX{} hyphenation
widespread in the computer world. That is a result of \TeX{} approach
simplicity and versatility.  The algorithm works effectively, as it already
supports rules for 66 languages \cite{texhyphenrules}, and offers the
flexibility to create rules for any currently unsupported languages.

Unfortunately, certain hyphenation rules cannot be implemented using the \TeX{}
hyphenation algorithm. In languages like German, hyphenation can lead to letter
changes or insertions. Additionally, compound words lack hyphens, resulting in
extended letter sequences without visible separation and even repetitions of
the same letter, as seen in examples like \emph{Wasserrinne} and
\emph{Schifffahrt}.  Furthermore, the German spelling reform made some changes,
making it necessary to create a different set of rules for German hyphenation.
For example, the word \emph{Schiffahrt} should be hyphenated as
\emph{Schiff-fahrt}, preserving the \emph{f}s from each word that makes this
compound. The hyphenation should insert a \emph{f} that is not part of the
written form. That was not a problem for the old written form of the word:
\emph{Schifffahrt}. Also, the old hyphenating rules of German grammar stated
the hyphenation \emph{Bäk-ker} for the word \emph{Bäcker}, \emph{Zuk-ker} for
the word \emph{Zucker} and \emph{pak-ken} for the word \emph{packen}. Now those
words are hyphenated as \emph{Bä-cker}, \emph{Zu-cker} and \emph{pa-cken},
respectively. Some words have also changed their hyphenation point after the
spelling reform. For example, \emph{Fen-ster} became \emph{Fens-ter} and
\emph{mei-stens} became \emph{meis-tens}. Some problems in compound word
hyphenation in \TeX{} are discussed in \cite{sojka1995a}.

@@

\section{PATGEN}\label{sec-patgen}

PATGEN utilizes a list of hyphenated words to define rules at various levels
and lengths, based on the patterns found in the data. It starts with short
patterns and incrementally increases their length until reaching the maximum
pattern length allowed by the user.  The objective is to keep the patterns as
concise as possible, as this enhances their generalizability.  As it advances
and incorporates longer patterns, PATGEN establishes exceptions.  In certain
cases, analyzing long patterns may be necessary, as some hyphenation points
could depend on characters far away from the breaking point\footnote{Some
    examples of hyphenation dependency on characters far from the break point:
    \emph{dem-o-crat} and \emph{de-moc-ra-cy}; \emph{as-pi-rin} and
    \emph{aspir-ing}; \emph{de-mon-stra-tive} and \emph{dem-on-stra-tion}.
}.

PATEGEN works on glyph\footnote{\emph{Glyph} is commonly used in linguistics,
typography, and computer graphics to refer to a specific graphical
representation of a character or symbol, which can be the entire symbol or a
distinct visual element within it.} indices rather than character codes. Each
glyph is represented by a single byte. That amount to 256 indices, where 13 of
them are reserved for the digits 0-9 and the characters `.', `-', and `*'. The
remaining 243 are used to represent symbols of the given language.  To run
PATGEN, a translation file is necessary. This file defines the values of
certain language-specific parameters (in the first line) and enumerates the
various forms in which language symbols may appear (all subsequent lines).  In
the first line, positions 1 and 2 are used to set the value of
\verb|lefthyphenmin|, and positions 3 and 4 are used to set the value of
\verb|righthyphenmin|.  These values determine the minimum length of a string
that may be generated by a hyphenation procedure. To set a single-digit value,
leave the first position blank, i.e., place a space in position 1 and 3 for
\verb|lefthyphenmin| and \verb|righthyphenmin|, respectively.  Positions 5, 6
and 7 are used to define alternative values for the special characters `.', `-'
and `*'.  The following lines use a delimiter to enclose each 'letter' of the
desired language alphabet, including its alternative representations.  The
first position of the line defines the delimiter, and each symbol of the
language can occupy as many positions as necessary, as long as the reserved
value for the delimiter is not used in the symbol's definition.  Consider the
following example for defining the letter `e' in Portuguese: 
\begin{verbatim}
XeXEXéXêXÉXÊX\'{e}X\^{e}X\'{E}X\^{E}XX 
\end{verbatim}
It represents the many ways in which the letter `e' might be found: lowercase,
uppercase, with or without acute or circumflex accents. Note that we have used
the direct input (using UTF-8 or other encoding that support the accented `e'
character) and also the compositional counterpart using the escape codes that
define each accent.  In this definition, we have assumed that the many forms in
which we may find the character \emph{e} will be equivalent for
hyphenation (pattern matching) purpose. As another example, see the next line
which defines the character $\pi$ (taken from \cite{haralambous2021}):
\begin{verbatim} 
#p#P#\varpi ## 
\end{verbatim}

The dictionary file is a list of pre-hyphenated words from which PATGEN
extracts patterns to create hyphenation rules. To ensure PATGEN's proper
functionality, the translation file and the dictionary file must utilize the
same encoding, even if it is a multi-byte encoding. The translation file
describes how to handle byte sequences representing a glyph, and PATGEN will
work seamlessly when there are at most 243 symbols in the given language.

%Those rules are organized
%in different levels, from 1 to 9, where odd numbers represent hyphenating
%levels and even numbers represent inhibiting levels. Each level works as an
%inhibiting level of it predecessor. 
%A pattern will consist of a string made of
%characters (from the language alphabet) possibly with a number in between,
%expressing the hyphenation/inhibition level and possibly word boundaries marker
%(the period) at the pattern edges. When there is no number between characters
%in a pattern, a zero is assumed, which means \emph{undefined} and no
%hyphenation point will be suggest at that location.

\section{\TeX{} hyphenation rules for portuguese}\label{sec-tex-hyphen-pt}
\cite{rezende1987} was the first to create patterns for Portuguese hyphenation
in \TeX{}. The pattern set was last updated in 2015, incorporating
contributions from José João Dias Almeida \cite{hyphpt}, resulting in a total
of 307 rules\footnote{The set of rules for Portuguese hyphenation is short when
compared to rules in other languages. For example, English currently has 4938
rules, Russian has 7023 rules and German has 34011 rules.}. 
These rules effectively hyphenate the majority of Portuguese words.
% - detalhar as regras de Rezende e Dias 
% - contabilizar a quantidade de palavras que não são contempladas pela proposta do autor
% - analisar a necessidade de excluir alguma regra (ao invés de apenas acrescentar novas)

However, in light of certain cases, we propose an analysis of the default rules
to identify areas for improvement. By addressing specific issues and
considering non-typical patterns, we aim to enhance hyphenation accuracy. The
methodology used for this analysis is described in Section
\ref{sec-methodology}.

\section{Methodology}\label{sec-methodology}
To assess the performance of each set of rules, we require a list of correctly
hyphenated words in Portuguese. 
We utilized word frequency data to avoid selecting excessively rare words, creating a minimal representative set. 
A comprehensive list was compiled using data from online dictionaries and
Portuguese corpora. Additionally, we referred to Portuguese grammars to
validate hyphenation (\emph{translineation}) rules and manually corrected the data as
needed.

@@

Initially, we selected
\href{https://www.linguateca.pt/cetenfolha/index_info.html}{CETENFolha}
CETENFolha as the source corpus, but this approach presented two issues: many
words were still missing, and the corpus was based on text from 1994, prior to
the implementation of the Orthographic Agreement.  The Orthographic Agreement
of the Portuguese Language was conducted in 1990, and its transition period
started in 2009, becoming mandatory in 2016. However, even after the agreement,
some words continue to have different spellings in the participating countries.
To address these idiosyncrasies, we aim to develop hyphenation rules that
accommodate variations in spelling. For instance, the word \emph{reception}
might be written as \emph{receção} (Portugal) or \emph{recepção} (Brazil); the
word \emph{action} might be written as \emph{acção} (Portugal) or \emph{ação}
(Brazil); and the word \emph{project} might be written as \emph{projecto}
(Portugal) or \emph{projeto} (Brazil).

@@

O conjunto de \NumberOfNewRules{} regras é apresentado a seguir:
\begin{multicols}{6}
    \footnotesize
    \verbatiminput{../data/patch.TeX.pt-br.patterns}
\end{multicols}
